DCMIP2016, Part 1: Models and Equation Sets

### 2017-04-13 DCMIP2016-Part1-Models.tex
- Cleaned up section 2 by tabulating important model properties, moved equation derivations to an appendix, and began framing a section on horizontal discretization

This paper is the first in a series of papers documenting the results of this workshop.  This paper is intended to provide a review of the multitude of technologies and techniques that have been developed for non-hydrostatic global atmospheric modeling and provide an introduction to the intercomparison strategy that has been employed by DCMIP.  Section \ref{sec:Notation} describes the unified notation that our project has adopted that will be relevant throughout all papers in this series.  Section \ref{sec:DynamicalCores} then provides a brief overview of each of the participating models.  The middle of this paper is dedicated to an overview of techniques available for building the infrastructure of a global dynamical core:  Section \ref{sec:ModelGrids} describes model grids that have been employed by modern modeling systems; section \ref{sec:EquationSets} describes different formulations of the moist dynamical equations; section \ref{sec:DiffusionStabilization} describes diffusion, stabilization, filters and fixers employed by these models; and section \ref{sec:TemporalDiscretizations} describes temporal discretizations.  Conclusions and discussion then follow in section \ref{sec:Conclusions}.

\begin{table}[h]
\caption{List of symbols used by DCMIP} \label{tab:symbols}
\begin{center}
\begin{tabular}{cl}
\hline Symbol & Description \\ \hline 
$\lambda$ & Longitude (in radians) \\
$\varphi$ & Latitude (in radians) \\
$z$ & Height with respect to mean sea level (set to zero) \\
$p_s$ & Surface pressure ($p_s$ of moist air if $q>0$) \\
$\Phi_s$ & Surface geopotential \\
$z_s$ & Surface elevation with respect to mean sea level (set to zero) \\
$u$ & Zonal wind \\
$v$ & Meridional wind \\
$w$ & Vertical velocity \\
$\omega$ & Vertical pressure velocity  \\
$D$ & Divergence\\
$\zeta$ & Vertical component of relative vorticity\\
$p$ & Pressure (pressure of moist air if $q>0$) \\
$\rho$ & Total air density \\
$\rho_d$ & Dry air density \\
$T$ &Temperature \\
$T_v$ & Virtual temperature \\
$\theta$ & Potential temperature \\
$\theta_v$ & Virtual potential temperature \\
$q$ & Specific humidity \\
$P_{ls}$ & Large-scale precipitation rate \\
$q$ & Water vapor mixing ratio \\
$q_c$ & Cloud water mixing ratio \\
$q_r$ & Rain water mixing ratio \\
$q_i$ & General tracer mixing ratio \\
\hline 
\end{tabular}
\end{center}
\end{table}

@@

This section provide a short description of unique features or design specifications from several modern dynamical cores.  A tabulation of the dynamical cores that have submitted simulation results and model descriptions can be found in Table \ref{tab:Models}.

\begin{table}[h]
\caption{List of dynamical cores participating in DCMIP which have submitted simulation results.} \label{tab:Models}
\begin{center}
\begin{tabular}{cll}
\hline Acronym & Model Name & Modeling Center or Group \\ \hline 
CSU & Colorado State University Model & Colorado State University \\
DYNAMICO & DYNAMICO & Institut Pierre Simon Laplace (IPSL), France \\
FV$^3$ & GFDL Finite-Volume Cubed-Sphere Dynamical Core & Geophysical Fluid Dynamics Laboratory \\
FVM & Finite Volume Module of the Integrated Forecasting System & European Centre for Medium-Range Weather Forecasts \\
GEM & Global Environmental Multiscale model & Environment and Climate Change Canada \\
HOMME & High-Order Method Modeling Environment & Sandia National Laboratories / University of Colorado, Boulder \\
ICON & Icosahedral Non-hydrostatic model & Max-Planck-Institut f\"ur Meteorologie \\
MPAS & Model for Prediction Across Scales & National Center for Atmospheric Research \\
NICAM & Non-hydrostatic Icosahedral Atmospheric Model & AORI / JAMSTEC / AICS, Japan \\
Tempest & Tempest Non-hydrostatic Atmospheric Model & University of California, Davis \\
\hline 
\end{tabular}
\end{center}
\end{table}

@@

The CSU model uses an optimized geodesic grid to discretize the sphere, with height as the vertical coordinate. The model is based on the non-hydrostatic Unified System of equations proposed by \cite{arakawa2009unification}, which filters vertically propagating sound waves but allows the Lamb wave and does not require a reference state. The horizontal wind field is determined by predicting the vertical component of the vorticity and the divergence of the horizontal wind, and then solving a pair of two-dimensional Poisson equations for a stream function and velocity potential. Time-differencing is based on the third-order Adams-Bashforth scheme. Horizontal diffusion is included in the form of a $\nabla^4_z()$ operator acting on the vorticity, divergence, potential temperature, and tracer.  
\subsection{DYNAMICO}

DYNAMICO is a mimetic finite-difference / finite-volume model, solving initially the hydrostatic primitive equations and recently extended to solve the shallow-atmosphere fully-compressible equations. The grid is an hexagonal C-grid in the horizontal and a Lorenz grid in the vertical. DYNAMICO's design uniquely combines:
 \begin{itemize}
\item A representation of the prognostic and diagnostic fields following the ideas of discrete differential geometry,\citep{Dubos2015DYNAMICO10}
\item A novel Hamiltonian formulation of the equations of motion in non-Eulerian coordinates  \citep{Dubos2014Equations} which is imitated at the discrete level using building blocks from the litterature,\citep{thuburn2009numerical,ringler2010unified}, and yielding, {\color{red}to a certain extent [what does this mean?]}, an energy-conserving spatial discretization, and
\item a novel explicit-implicit splitting which results in a simple, efficient and scalable implicit solver while allowing stable time steps close or identical to those of the hydrostatic solver (Dubos and Dubey, in preparation). 
\end{itemize} In addition, it features a conservative positive-definite transport scheme based on a slope-limited finite-volume approach \citep{Dubey2015Intercomparison}.

@@

The GFDL Finite-Volume Cubed-Sphere Dynamical Core (FV$^3$, or sometimes written FV3) is a fully finite-volume discretization of the fully-compressible non-hydrostatic Euler equations. The discretization is done in the horizontal on the equiangular gnomonic cubed-sphere grid \citep{PL2007JCP}, and a flow-following Lagrangian vertical coordinate \citep{lin2004vertically}. The Lagrangian vertical coordinate deforms so that the flow is constrained to follow the Lagrangian surfaces, allowing vertical transport to be represented implicitly without additional advection terms (see section \ref{sec:FloatingLagrangian} below). Fluxes are computed using the Piecewise-Parabolic Method of \citet{CW1984JCP} with an optional monotonicity constraint; in non-hydrostatic applications the monotonicity constraint is used primarily for tracer transport. The discretization is on the C-D grid as described by \citet{LR1997QJR} which acts as a simplified Riemann solver: the D-grid winds are interpolated to the C-grid and then advanced by half of an acoustic timestep, giving time-centered winds
that can then be used to compute the fluxes and advance the flux terms by a full acoustic timestep. Since divergence is effectively invisible to the solver, a divergence damping is applied to control numerical noise as divergent modes cascade to the grid scale.

Implicit viscosity is applied through the monotonicity constraint; if non-monotonic advection is used for the momentum and total air mass a weak explicit hyperviscosity is applied for stability and to alleviate numerical noise. Explicit viscosity is applied every acoustic timestep.

The prognostic variables are horizontal winds, in the native Gnomonic local coordinate; virtual potential temperature, which is conserved by the adiabatic dynamics; mass, represented by the difference in hydrostatic pressure between the top and bottom of a grid cell; and tracer mass. The non-hydrostatic solver adds a prognostic vertical velocity and geometric height of each grid cell, which can then be used to compute density. All variables are 3D cell-mean values, except for the horizontal winds, which are 2D face-mean values on their respective staggerings; as a result, vorticity is a 3D cell-mean value.

@@

The finite-volume module (FVM) of the Integrated Forecasting System (IFS) is presently under development at 
ECMWF \citep{smolarkiewiczetalJCP2016}. FVM solves the fully compressible Euler 
equations in geospherical coordinates. Both deep-atmosphere and shallow-atmosphere equations 
are available by means of simple switches. The formulation incorporates 
a generalized, optionally time-dependent, terrain-following vertical coordinate based on height. 
A centered two-time-level semi-implicit integration scheme is employed with 3D implicit treatment 
of acoustic, buoyant, and rotational modes \citep{smolarkiewiczJCP2014}. The associated 3D 
Helmholtz problem is solved iteratively using a bespoke preconditioned Generalised 
Conjugate Residual approach. 
The integration procedure uses the multidimensional flux-form Eulerian non-oscillatory MPDATA 
advection scheme \citep{smolarkiewiczszmelter2005,kuehnleinJCP2017}. 
The horizontal spatial discretization is fully unstructured finite-volume using the 
median-dual approach. This is combined with a structured-grid finite-difference approach in the 
vertical direction; see \cite{smolarkiewiczetalJCP2016} for an exposition. 
In both the horizontal and the vertical discretization, all prognostic variables are 
co-located. The median-dual finite-volume mesh in the horizontal is developed about the points/nodes 
of the octahedral reduced Gaussian grid (Section~\ref{subsection:octahedralreducedgaussiangrid}). The octahedral 
reduced Gaussian grid is also employed in the spectral dynamical core of the current operational 
IFS at ECMWF, which facilitates interoperability of the two formulations. However, we note that FVM is 
not restricted to this grid and offers capabilities towards a broad classes of meshes including adaptivity.

No explicit diffusion is applied in FVM for DCMIP, apart from the momentum dissipation and scalar 
diffusion required for some of the test cases, which is the vertical dissipation/diffusion in the planetary
boundary layer parametrization and the constant-coefficient second-order dissipation/diffusion in 
the supercell test. An absorbing layer in the first latitude ring around the poles is optionally used in 
the form of a Rayleigh-type forcing to the prognostic variables. The dynamics time step is adapted 
at every time step according to a given maximum advective CFL number (typically somewhat smaller 
than 1). The physics time step is identical to the dynamics time step. 
%The first row of the table below refers to the size of the octahedral reduced Gaussian grid, where 
%the number following "O" (for octahedral) denotes the number of latitudes between pole and equator.

@@

The GEM model \citep{Girard2014}, which is used for operational forecasting at Environment and Climate Change Canada, uses the Yin-Yang grid \citep{kageama2004yinyang} and horizontal discretization on an Arakawa C grid.
The vertical coordinate is a hybrid terrain-following coordinate of a log-hydrostatic-pressure type
and the vertical discretization is based on the Charney-Phillips grid.
A two time level semi-Lagrangian implicit time discretization is implemented as described in section \ref{sec:GEM_temporal}.
It gives rise to an {\it{iterative}} process where each step requires the solution of a
linear system of equations that is reduced to a Helmholtz problem for one composite variable.
For this problem, a direct solver is involved, using the Schwarz-type domain decomposition method on a Yin-Yang grid \citep{Qaddouri2008schwarz}.
The dynamics and physics are time split.
An explicit hyperviscosity is employed for wind components and tracers
via applications of the Laplacian operator.
Viscosity operations are applied after the completion of the physics time step.

@@

The High Order Method Modeling Environment (HOMME) \citep{taylor2010compatible, dennis2012cam} is currently used by both the Community Atmosphere Model Spectral Element (CAM-SE) dynamical core and Accelerated Climate Model for Energy Atmosphere (ACME-A) dynamical core.  As a general framework for several dynamical methods, HOMME includes a hydrostatic option and an experimental non-hydrostatic core, although only the shallow-atmosphere are available at present. The model is designed to be mass and energy conserving, with nearly optimal parallel scalability at large core counts. The CAM-SE dynamical core is a hydrostatic model that partitions the globe horizontally using an unstructured grid of quadrilateral elements. These elements are arranged by default in a cubed-sphere structure, but arbitrary quadrilateral grids with conforming edges may also be employed.  The model employs a hybrid terrain-following / pressure coordinate. The fluid equations are discretized using dimension splitting, with a nodal 4th order spectral element discretization in the horizontal and the mimetic (mass and energy conserving) 2nd order finite difference discretization of \cite{simmons1981energy} in the vertical. Fields are co-located in the horizontal in the sense that they share the same 4th order basis functions. Lorenz staggering is employed in the vertical with (u,v,T,p) placed on layer midpoints while vertical fluxes $m \dot \eta$ are placed on layer interfaces.

@@

The ICON model {\color{red}[Reference?]} discretizes the compressible equations for a shallow atmosphere in vector invariant form for the horizontal wind on a triangular Arakawa C-grid and a smoothed terrain following height based Lorenz grid. The prognostic variables are the normal wind $v_n$ at the edge mid points of full levels, the vertical wind $w$ in the circumcenters of the triangles on half levels and virtual potential temperature $\theta_v$, full air density $\rho$, which includes moisture and hydrometeor densities and tracer mixing ratios $q_x$ with respect to the full air density. The discretization in time employs a two time level predictor corrector scheme, which is explicit in all terms except for those describing the vertical propagation of sound waves. Time splitting is applied between the dynamics that is forced by slow physics on the one hand and horizontal diffusion, tracer transport, and fast physics. One complete time step typically includes 5 dynamical sub-steps. The average air mass flux of the dynamical  sub-steps is provided to the tracer transport to allow for a mass-consistent transport. For stabilization of the divergence term on the triangular C-grid the divergence in a triangle is computed from modified normal wind components resulting from a weighted average including normal winds on edges of adjacent cells. Further divergence damping is applied to the normal wind at every sub-step. Rayleigh damping is applied to the vertical wind in layers close to the model top in order to avoid the reflection of gravity waves. The horizontal diffusion, which is applied at full model time steps, combines a flow dependent Smagorinski scheme with a background 4th order Laplacian diffusion operator. For tracer transport a flux form semi-Lagrangian scheme with monotone flux limiters is used, which grants local mass conservation and consistency with the air motion. The numerical methods have been chosen for high numerical efficiency, and they rely on next neighbour communication only, thus allowing massive parallelization.

@@

The Model for Prediction Across Scales (MPAS) \citep{skamarock2012multiscale} uses an Arakawa C-grid built on a centroidal Voronoi tessellation, in conjunction with the mimetic TRiSK discretization \citep{thuburn2009numerical,ringler2010unified}.  Advection terms are nominally third- to fourth-order and are handled in accordance with \cite{skamarock2011conservative}.  In the vertical, MPAS employs a Lorenz-type second-order nodal finite volume method with a smoothed terrain-following height coordinate.  The prognostic variables are dry air pseudodensity $\tilde{\rho}_d$, dry momentum $\tilde{\rho}_d \vb{u}$, and a modified moist potential temperature.  Integration in time is handled via the split-explicit method of \cite{klemp2007conservative}.  Various filters are available for controlling spurious oscillations, including Smagorinsky-type eddy viscosity, and fourth-order hyperdiffusion.

>>>>

This paper is the first in a series of papers documenting the results of this workshop.  This paper is intended to provide a review of the multitude of technologies and techniques that have been developed for non-hydrostatic global atmospheric modeling and provide an introduction to the intercomparison strategy that has been employed by DCMIP.  For easy reference, a list of mathematical symbols that are employed in this paper (and subsequent DCMIP papers) is given in Table \ref{tab:symbols}.  Section \ref{sec:DynamicalCores} provides a brief overview of each of the participating models.  The middle of this paper is dedicated to an overview of techniques available for building the infrastructure of a global dynamical core:  Section \ref{sec:ModelGrids} describes model grids that have been employed by modern modeling systems; section \ref{sec:EquationSets} describes different formulations of the moist dynamical equations; section \ref{sec:DiffusionStabilization} describes diffusion, stabilization, filters and fixers employed by these models; and section \ref{sec:TemporalDiscretizations} describes temporal discretizations.  Conclusions and discussion then follow in section \ref{sec:Conclusions}.

\begin{table}[h]
\caption{List of symbols used by DCMIP} \label{tab:symbols}
\begin{center}
\begin{tabular}{cl}
\hline Symbol & Description \\ \hline 
$\lambda$ & Longitude (in radians) \\
$\varphi$ & Latitude (in radians) \\
$z$ & Height with respect to mean sea level (set to zero) \\
$p_s$ & Surface pressure ($p_s$ of moist air if $q>0$) \\
$\Phi_s$ & Surface geopotential \\
$z_s$ & Surface elevation with respect to mean sea level (set to zero) \\
$u$ & Zonal wind \\
$v$ & Meridional wind \\
$w$ & Vertical velocity \\
$\vb{u}$ & 3D wind vector \\
$\vb{u}_h$ & Horizontal wind vector \\
$\vb{v}_h$ & Horizontal wind vector with covariant components \\
$\omega$ & Vertical pressure velocity  \\
$D$ & Divergence of the horizontal wind vector\\
$\zeta$ & Vertical component of relative vorticity\\
$p$ & Pressure (pressure of moist air if $q>0$) \\
$e$ & Internal energy \\
$\rho$ & Total air density \\
$\rho_d$ & Dry air density \\
$\rho_s$ & Pseudo-density \\
$T$ &Temperature \\
$T_v$ & Virtual temperature \\
$\theta$ & Potential temperature \\
$\theta_v$ & Virtual potential temperature \\
$q$ & Specific humidity \\
$P_{ls}$ & Large-scale precipitation rate \\
$q$ & Water vapor mixing ratio \\
$q_c$ & Cloud water mixing ratio \\
$q_r$ & Rain water mixing ratio \\
$q_i$ & General tracer mixing ratio \\
\hline 
\end{tabular}
\end{center}
\end{table}

@@

This section provide a brief overview of key discretization choices, along with unique features or design specifications from participating dynamical cores.  Further details on these choices can be found in subsequent sections.  Simulation results and model descriptions have been submitted from ten dynamical cores (see Table \ref{tab:Models}).  Although all models solve the Euler equations (or a close approximation of these equations), essentially every model employs a different formulation with a distinct set of prognostic variables.  {\color{red}[Maybe move to a separate paper?] The derivation of many of these formulations can be found in Appendix \ref{sec:EquationSets}.}  The particular choice of equation set,  that have been made by these models.

A tabulation of the dynamical cores that have submitted simulation results and model descriptions can be found in Table \ref{tab:Models}.

\begin{table}[h]
\caption{Participating modeling centers and associated dynamical cores that have submitted simulation results.} \label{tab:Models}
\begin{center}
\begin{tabular}{cll}
\hline Acronym & Model Name & Modeling Center or Group \\ \hline 
CSU & Colorado State University Model & Colorado State University, USA \\
DYNAMICO & DYNAMICO & Institut Pierre Simon Laplace (IPSL), France \\
FV$^3$ & GFDL Finite-Volume Cubed-Sphere Dynamical Core & Geophysical Fluid Dynamics Laboratory, USA \\
FVM & Finite Volume Module of the Integrated Forecasting System & European Centre for Medium-Range Weather Forecasts \\
GEM & Global Environmental Multiscale model & Environment and Climate Change Canada \\
HOMME & High-Order Method Modeling Environment & Sandia National Laboratories / University of Colorado, Boulder \\
ICON & Icosahedral Non-hydrostatic model & Max-Planck-Institut f\"ur Meteorologie, Germany \\
MPAS & Model for Prediction Across Scales & National Center for Atmospheric Research, USA \\
NICAM & Non-hydrostatic Icosahedral Atmospheric Model & AORI / JAMSTEC / AICS, Japan \\
Tempest & Tempest Non-hydrostatic Atmospheric Model & University of California, Davis, USA \\
\hline 
\end{tabular}
\end{center}
\end{table}

\begin{table}[h]
\caption{Details on the prognostic variables and horizontal discretization for participating dynamical cores.  Equation set indicates whether a model is hydrostatic (H) or nonhydrostatic (NH), and whether the model presently supports the deep-atmosphere formulation (D).  Only two numerical methods are represented among participating models, namely finite-volume (FV) and spectral-element (SE).  More details on horizontal staggering can be found in section \ref{sec:HorizontalStaggering}.} \label{tab:ModelsHorizontal}
\begin{center}
\begin{tabular}{cccccc}
\hline Acronym & Equation set & Prognostic Variables & Horizontal Grid & Numerical & Horizontal  \\
& & & & method & staggering \\ \hline
CSU & NH (Unified) & $\zeta$, $D$, $w$, $p_s$, $\theta_v$, $q$ & Geodesic (\S \ref{sec:grid_geo}) & FV & Z-grid \\
DYNAMICO & H/NH {\color{red}(D)} & $\vb{v}_h$, $\rho_s w$, $\rho_s$, $\rho_s \theta_v$, $\Phi$, $\rho_s q$ & Geodesic (\S \ref{sec:grid_geo}) & FV & C-grid \\
FV$^3$ & NH & $\vb{u}_h$, $w$, $\rho_s$, $\rho_s \theta_v$, $\Phi$, $\rho_s q$ & Cubed-sphere (\S \ref{sec:grid_cs}) & FV & D-grid \\
FVM & NH (D) & $\rho_d$, $\rho_d \vb{u}_h$, $\rho_d w$, $\rho_d \theta$, ($\Phi$), $\rho_d q$ & Octahedral reduced (\S \ref{sec:grid_oct}) & FV & A-grid \\
GEM & NH & $\vb{u}_h$, $w$, $\log \partial s / \partial \zeta$, $\log T_v$, $\Phi$, $q$ & Yin-yang (\S \ref{sec:grid_yinyang}) & FV & C-grid  \\
HOMME & H & $u$, $v$, $\rho_s$, $T_v$, $\rho_s q$ & Cubed-sphere (\S \ref{sec:grid_cs}) & SE & A-grid \\
ICON & NH & $\vb{u}_h$, $w$, $\rho$, $\theta_v$, $\rho q$ & Icosahedral triangular (\S \ref{sec:grid_ico_tri}) & FV & C-grid \\
MPAS & NH & $\rho_d \vb{u}_h$, $\rho_d w$, $\rho_d$, $\rho_d \theta_v$, $\rho_d q$ & CCVT (\S \ref{sec:grid_ccvt}) & FV & C-grid \\
NICAM & NH & $\rho \vb{u}_h$, $\rho w$, $\rho$, $\rho e$, $\rho q$ & Geodesic (\S \ref{sec:grid_geo}) & FV & A-grid \\
Tempest & NH & $\vb{u}_h$, $w$, $\rho$, $\rho \theta_v$, $\rho q$ & Cubed-sphere (\S \ref{sec:grid_cs}) & SE & A-grid \\
\hline 
\end{tabular}
\end{center}
\end{table}

\begin{table}[h]
\caption{Vertical coordinates and staggering for participating dynamical cores.  More details on vertical staggerings can be found in section \ref{sec:VerticalStaggering}.} \label{tab:ModelsVertical}
\begin{center}
\begin{tabular}{cccccc}
\hline Acronym & Vertical Coordinate & Vertical Staggering \\ \hline 
CSU & fixed height & Lorenz \\
DYNAMICO & {\color{red}floating mass?} & Lorenz \\
FV$^3$ & floating mass & Lorenz \\
FVM & (floating) height & Co-located \\
GEM & log pressure & Modified Charney-Phillips (\S \ref{sec:VerticalStaggering})  \\
HOMME & floating mass & Co-located \\
ICON & fixed height & Lorenz \\
MPAS & fixed height & Lorenz \\
NICAM & fixed height & Lorenz \\
Tempest & fixed height & Lorenz \\
\hline 
\end{tabular}
\end{center}
\end{table}

\begin{table}[h]
\caption{Standard options for diffusion, filters, or fixers in participating dynamical cores, detailed in section \ref{sec:DiffusionStabilization}, and temporal discretization, detailed in section \ref{sec:TemporalDiscretizations}.} \label{tab:ModelsTemporalDiffusion}
\begin{center}
\begin{tabular}{cccccc}
\hline Acronym & Standard options for diffusion, & Temporal discretization \\
& stabilization, filters, or fixers & \\ \hline 
CSU & 4th-order horizontal hyperdiffusion & 3rd-order Adams-Bashforth (AB3) \\
DYNAMICO & {\color{red}?} & ARK232 \citep{giraldo2013implicit} \\
FV$^3$ & Divergence damping, monotonic limiting, hyperviscosity & Forward-backwards \citep{LR1997QJR} \\
FVM & Momentum dissipation, scalar diffusion, absorbing layer & Semi-implicit \citep{smolarkiewiczJCP2014} (\S \ref{sec:FVMSemiImplicit}) \\
GEM & Hyperviscosity & Semi-implicit \citep{Girard2014} (\S \ref{sec:GEM_temporal}) \\
HOMME & Hyperdiffusion & KGU53 \citep{guerra2016high} \\
ICON & Divergence damping, Smagorinsky, hyperdiffusion & Predictor-corrector \\
MPAS & Smagorinsky, hyperdiffusion & Split-explicit \citep{klemp2007conservative} \\
NICAM & 3D divergence damping, Smagornisky, hyperviscosity & Split-explicit \citep{klemp2007conservative} \\
Tempest & Hyperdiffusion & ARS232 \citep{ascher1997implicit} \\
\hline 
\end{tabular}
\end{center}
\end{table}

@@

The CSU model uses an optimized geodesic grid to discretize the sphere, with height as the vertical coordinate. The model is based on the non-hydrostatic Unified System of equations proposed by \cite{arakawa2009unification}, which filters vertically propagating sound waves but allows the Lamb wave and does not require a reference state. The horizontal wind field is determined by predicting the vertical component of the vorticity and the divergence of the horizontal wind, and then solving a pair of two-dimensional Poisson equations for a stream function and velocity potential. Horizontal diffusion is included in the form of a fourth-order Laplacian operator applied on constant height surfaces ($\nabla^4_z$) that acts on the vorticity, divergence, potential temperature, and tracer.  The CSU model supports both third-order and fifth-order upstream-weighted finite-volume advection schemes, with positivity preservation enforced via mass borrowing.
\subsection{DYNAMICO}

DYNAMICO is a mimetic finite-difference / finite-volume model, solving initially the hydrostatic primitive equations and recently extended to solve the shallow-atmosphere fully-compressible equations. DYNAMICO's design uniquely combines a representation of the prognostic and diagnostic fields following the ideas of discrete differential geometry, \citep{Dubos2015DYNAMICO10}.  It includes a novel Hamiltonian formulation of the equations of motion in non-Eulerian coordinates \citep{Dubos2014Equations} which is imitated at the discrete level using building blocks from the literature,\citep{thuburn2009numerical,ringler2010unified}, and (up to the addition of explicit diffusion) leads to an energy-conserving spatial discretization.  It also incorporates a novel explicit-implicit splitting which results in a simple, efficient and scalable implicit solver while allowing stable time steps close or identical to those of the hydrostatic solver (Dubos and Dubey, in preparation).  In addition, it features a conservative positive-definite transport scheme based on a slope-limited finite-volume approach \citep{Dubey2015Intercomparison}.

@@

The GFDL Finite-Volume Cubed-Sphere Dynamical Core (FV$^3$, or sometimes written FV3) is a fully finite-volume discretization of the fully-compressible non-hydrostatic Euler equations on the equiangular gnomonic cubed-sphere grid with flow-following Lagrangian vertical coordinate \citep{lin2004vertically}. The Lagrangian vertical coordinate deforms so that the flow is constrained to follow the Lagrangian surfaces, allowing vertical transport to be represented implicitly without additional advection terms (see section \ref{sec:FloatingLagrangian} below). Fluxes are computed using the Piecewise-Parabolic Method of \citet{CW1984JCP} with an optional monotonicity constraint; in non-hydrostatic applications the monotonicity constraint is used primarily for tracer transport. The discretization is on the C-D grid as described by \citet{LR1997QJR} (see section \ref{sec:CDGrid}). Since divergence is effectively invisible to the solver, a divergence damping is applied to control numerical noise as divergent modes cascade to the grid scale.  Implicit viscosity is applied through the monotonicity constraint; if non-monotonic advection is used for the momentum and total air mass a weak explicit hyperviscosity is applied for stability and to alleviate numerical noise. Explicit viscosity is applied every acoustic timestep.

The prognostic horizontal winds are stored in the native Gnomonic local coordinate. The non-hydrostatic solver adds a prognostic vertical velocity and geometric height of each grid cell, which can then be used to compute density. All variables are 3D cell-mean values, except for the horizontal winds, which are 2D face-mean values on their respective staggerings; as a result, diagnostic vorticity is a 3D cell-mean value.

@@

The finite-volume module (FVM) of the Integrated Forecasting System (IFS) is presently under development at 
ECMWF \citep{smolarkiewiczetalJCP2016}. FVM solves the fully compressible Euler 
equations in geospherical coordinates. Both deep-atmosphere and shallow-atmosphere equations 
are available by means of simple switches. The formulation incorporates 
a generalized, optionally time-dependent, terrain-following vertical coordinate based on height. 
A centered two-time-level semi-implicit integration scheme is employed with 3D implicit treatment 
of acoustic, buoyant, and rotational modes \citep{smolarkiewiczJCP2014} (see section \ref{sec:FVMSemiImplicit}). The associated 3D 
Helmholtz problem is solved iteratively using a bespoke preconditioned Generalised 
Conjugate Residual approach. The dynamics time step is adapted 
at every time step according to a given maximum advective CFL number (typically somewhat smaller 
than 1). The physics time step is identical to the dynamics time step.

The integration procedure uses the multidimensional flux-form Eulerian non-oscillatory MPDATA 
advection scheme \citep{smolarkiewiczszmelter2005,kuehnleinJCP2017}. 
The horizontal spatial discretization is fully unstructured finite-volume using the 
median-dual approach. This is combined with a structured-grid finite-difference approach in the 
vertical direction; see \cite{smolarkiewiczetalJCP2016} for an exposition. 
In both the horizontal and the vertical discretization, all prognostic variables are 
co-located. The median-dual finite-volume mesh in the horizontal is developed about the points/nodes 
of the octahedral reduced Gaussian grid (see section \ref{sec:grid_oct}). The octahedral 
reduced Gaussian grid is also employed in the spectral dynamical core of the current operational 
IFS at ECMWF, which facilitates interoperability of the two formulations. However, we note that FVM is 
not restricted to this grid and offers capabilities towards a broad classes of meshes including adaptivity.

No explicit diffusion is applied in FVM for DCMIP, apart from the momentum dissipation and scalar 
diffusion required for some of the test cases, which is the vertical dissipation/diffusion in the planetary
boundary layer parametrization and the constant-coefficient second-order dissipation/diffusion in 
the supercell test. An absorbing layer in the first latitude ring around the poles is optionally used in 
the form of a Rayleigh-type forcing to the prognostic variables.

%The first row of the table below refers to the size of the octahedral reduced Gaussian grid, where 
%the number following "O" (for octahedral) denotes the number of latitudes between pole and equator.

@@

The GEM model \citep{Girard2014} is used for operational forecasting at Environment and Climate Change Canada.  The horizontal discretization uses the Yin-Yang grid \citep{kageama2004yinyang}, with Arakawa C-grid staggering.
The vertical coordinate is a unique hybrid terrain-following coordinate of a log-hydrostatic-pressure type
and the vertical discretization is based on the Charney-Phillips grid (see section \ref{sec:VerticalStaggering}).
A two time level semi-Lagrangian implicit time discretization is implemented as described in section \ref{sec:GEM_temporal}.
It gives rise to an {\it{iterative}} process where each step requires the solution of a
linear system of equations that is reduced to a Helmholtz problem for one composite variable.
For this problem, a direct solver is involved, using the Schwarz-type domain decomposition method \citep{Qaddouri2008schwarz}.
The dynamics and physics are time split.
To eliminate numerical noise, an explicit hyperviscosity is employed for wind components and tracers
via applications of the Laplacian operator.
Viscosity operations are applied after the completion of the physics time step.

@@

The High Order Method Modeling Environment (HOMME) \citep{taylor2010compatible, dennis2012cam} is currently used by both the Community Atmosphere Model Spectral Element (CAM-SE) dynamical core and Accelerated Climate Model for Energy Atmosphere (ACME-A) dynamical core.  As a general framework for several dynamical methods, HOMME includes a hydrostatic option and an experimental non-hydrostatic core, although only shallow-atmosphere mode is available at present. The model is designed to be mass and energy conserving, with nearly optimal parallel scalability at large core counts. The CAM-SE dynamical core is a hydrostatic model that partitions the globe horizontally using an unstructured grid of quadrilateral elements. These elements are arranged in a cubed-sphere structure, although arbitrary quadrilateral grids with conforming edges may also be employed.  The model employs a hybrid terrain-following pressure coordinate in the vertical. The fluid equations are discretized using dimension splitting, with a nodal 4th order spectral element discretization in the horizontal and the mimetic (mass and energy conserving) 2nd order finite difference discretization of \cite{simmons1981energy} in the vertical. Fields are co-located in the horizontal in the sense that they share the same 4th order basis functions. Lorenz staggering is employed in the vertical with (u,v,T$_v$,p) placed on layer midpoints and diagnostic vertical fluxes $m \dot \eta$ are computed on layer interfaces.

@@

The ICON model \citep{zangl2015icon} discretizes the compressible equations for a shallow atmosphere in vector invariant form for the horizontal wind on a triangular Arakawa C-grid and a smoothed terrain-following height-based Lorenz grid.  Prognostic horizontal velocities are stored as normal wind components $v_n$ at the edge mid-points of full levels.  Prognostic vertical wind $w$ is stored at the circumcenters of the triangles on half levels. The discretization in time employs a two time level predictor corrector scheme, which is explicit in all terms except for those describing the vertical propagation of sound waves. Time splitting is applied between the dynamics that is forced by slow physics on the one hand and horizontal diffusion, tracer transport, and fast physics. One complete time step typically includes 5 dynamical sub-steps. The average air mass flux of the dynamical  sub-steps is provided to the tracer transport to allow for a mass-consistent transport. For stabilization of the divergence term on the triangular C-grid the divergence in a triangle is computed from modified normal wind components resulting from a weighted average including normal winds on edges of adjacent cells. Further divergence damping is applied to the normal wind at every sub-step. Rayleigh damping is applied to the vertical wind in layers close to the model top in order to avoid the reflection of gravity waves. The horizontal diffusion, which is applied at full model time steps, combines a flow dependent Smagorinsky scheme with a background 4th order Laplacian diffusion operator. For tracer transport a flux form semi-Lagrangian scheme with monotone flux limiters is used, which grants local mass conservation and consistency with the air motion. The numerical methods have been chosen for high numerical efficiency, and they rely on next-neighbor communication only, thus allowing massive parallelization.

@@

The Model for Prediction Across Scales (MPAS) \citep{skamarock2012multiscale} uses an Arakawa C-grid built on a centroidal Voronoi tessellation (see section \ref{sec:grid_ccvt}), in conjunction with the mimetic TRiSK discretization \citep{thuburn2009numerical,ringler2010unified}.  Advection terms are nominally third- to fourth-order and are handled in accordance with \cite{skamarock2011conservative}.  In the vertical, MPAS employs a Lorenz-type second-order nodal finite volume method with a smoothed terrain-following height coordinate.  The prognostic variables are dry air pseudodensity $\tilde{\rho}_d$, dry momentum $\tilde{\rho}_d \vb{u}$, and a modified moist potential temperature.  Integration in time is handled via the split-explicit method of \cite{klemp2007conservative}.  Various filters are available for controlling spurious oscillations, including Smagorinsky-type eddy viscosity, and fourth-order hyperdiffusion.

\subsection{NICAM}

NICAM uses the finite-volume discretization of the fully compressible non-hydrostatic Euler equations on a geodesic grid optimized by the spring dynamics method proposed by \cite{tomita2002optimization}. The system of equations is transformed to the standard terrain-following height coordinate system \citep{tomita2004new}. Instead of the temperature or the potential temperature, the total energy is chosen as a prognostic variable, following the method of \citep{satoh2002conservative, satoh2003conservative}. All of the prognostic variables are collocated horizontally at the mass centroid of each hexagonal/pentagonal cell to mitigate accuracy reduction under cell averaging, which is required in converting cell integrated quantities to point values at cell centroids. The use of cell centroids ensures quasi second-order accuracy of the gradient and divergence operators of NICAM \citep{tomita2001shallow}. The Arakawa C-grid staggering is used vertically. A two-stage Runge-Kutta scheme is usually used to reduce computational burden although a three-stage Runge-Kutta scheme \citep{wicker2002time} is recommended. The split-explicit time discretization is used for the horizontally propagating sound waves with the divergence damping term \citep{skamarock1992stability}. The implicit time discretization is adopted for the vertically propagating one. A variant of the piecewise linear transport scheme \citep{miura2007upwind, niwa2011three} is used with a flux limiter of \cite{thuburn1997pv} for passive tracer transports.
